---
title: "Final Report: COVID-19 Case Rate vs Population Demographics and Mask Policy"
author: "Chen WANG, XUAN LI"
date: "2022-12-19"
output: html_document
---
reload the data
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lmtest)
library(sandwich)
library(tidyverse)
library(magrittr)
library(ggplot2)
library(patchwork)
library(stargazer)
library(tinytex)
library(dplyr)
library(car)
library(MASS)
library(corrplot)
#install.packages('scico')
library(scico)
#install.packages("kableExtra")
library(kableExtra)
setwd("C:/Users/wangchen/Desktop/897/final report")
getwd()
```

```{r variables, warning=FALSE, results='hide'}
data <- read.csv("data_psacalc.csv", header = TRUE)
data <- rename(data,c(adm_2 = administered.02.01,
                adm_4 = administered.04.01,
                adm_6 = administered.06.01,
                adm_8 = administered.08.01,
                adm_9 = administered.09.21,
                mask_md = No_legal_mask_mandate))
```

categorical variable counts:
```{r}
cat('Count of states that enforced a mask mandate: ', 
    length(data$mask_md[data$mask_md==1]))

cat('\nCount of states did not enforce a mask mandate: ', 
    length(data$mask_md[data$mask_md==0]))
```

```{r}
data_c <- na.omit(data[, c('state','adm_2',
                 'adm_4',
                 'adm_6',
                 'adm_8',
                 'adm_9',
                 'exp',
                 'phexp1',
                 'vaxref_jan',
                 'nodoc',
                 'dmargin',
                 'density',
                 'poverty',
                 'plus65',
                 'mask_md')])
```


```{r}
#Here, we look at monthly new vaccinations instead of cumulative vaccinations.
data_month <- na.omit(data[,c('state',"exp", 
                         "phexp1", 
                         "vaxref_jan", 
                         "nodoc", 
                         "dmargin", 
                         "density", 
                         "poverty", 
                         "plus65",
                         "mask_md")]) 
#Subtractions to obtain the monthly tallies
data_month$feb_adm <- data$administered.02.28 - data$adm_2
data_month$april_adm <- data$administered.04.30 - data$administered.03.31
data_month$june_adm <- data$administered.06.30 - data$administered.05.31
data_month$august_adm <- data$administered.08.31 - data$administered.07.31
#For september we take 21st as end of the month.
data_month$sept_adm <- data$adm_9 - data$administered.08.31

sum(is.na(data_month$feb_adm))
sum(is.na(data_month$april_adm))
sum(is.na(data_month$june_adm))
sum(is.na(data_month$august_adm))
sum(is.na(data_month$sept_adm))
```

Descriptive statistics
```{r}
dataR <- cbind(data_c,data_month)
dataR <- dataR[, !duplicated(colnames(dataR))]
dataR <- na.omit(dataR)
stargazer(dataR,
          type = 'html', 
          min.max=TRUE, mean.sd = TRUE, 
          nobs = TRUE, median = TRUE, iqr = FALSE,
          digits=2, align=T,
          title = "Table 1. Descriptive Statistics", 
          out='table.doc')
```

Correlation matrix
```{r}
#correlation matrix
cor_max= cor(data_c[,-c(1:3,5:6)], use="complete.obs", method="pearson")

stargazer(cor(cor_max), 
          title = "Table 2. variable correlations",
          type = "html",
          out='covariance.doc')
 
pdf("corrplot.pdf", width = 5, height = 5) 
corrplot(cor(data_c[,-c(1:3,5:6)]),  
         col = scico(200, direction = -1, palette = "batlow"), 
         tl.col = "black") 
dev.off() 
```

```{r}
#This section produces the cross-sectional regression tables for April 1, June 1, and August 1 with additional controls.

mod1 <- lm(adm_2 ~ exp + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65 + mask_md, data=dataR)

mod2 <- lm(adm_4 ~ exp + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65 + mask_md, data=dataR)

mod3 <- lm(adm_6 ~ exp + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65 + mask_md, data=dataR)

mod4 <- lm(adm_8 ~ exp + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65 + mask_md, data=dataR)

mod5 <- lm(adm_9 ~ exp + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65 + mask_md, data=dataR)

#examine the multicolinearity
kable(vif(mod3))

stargazer(mod2, mod3, mod4,
          title = 'Table 3. regression result', 
          ci=TRUE, ci.level = 0.95, 
          single.row =FALSE, 
          digits = 2, 
          no.space = TRUE, 
          type ='html',
          out = 'regression1.doc')

```
Scatterplot for linearity (2 ways of plots)
```{r}
# linearity
scatterplotMatrix(data_c[,-c(1:3,5:6,15)],
                  cex = .5,
                  pch = 16,
                  col = rgb(0,0,0,1/32),
                  diagonal = list(method = "histogram",
                                  breaks = 20),
                  cex.labels = 0.5,
                  regLine = list(method = lm,
                                 lty = 1,
                                 lwd = 1,
                                 col = 1),
                  smooth = list(method = "loessLine",
                                lty.smooth = 2,
                                lwd.smooth = 1,
                                col.smooth = 2,
                                lty.spread = 3,
                                lwd.spread = 1,
                                col.spread = 2),
                  main = "Figuer 2: Scatter Plot Matrix")
```

```{r}
scatterplot(adm_8 ~ exp,  
            data = data_c, 
            boxplots = F, 
            grid = F, 
            jitter = list(x = 1), 
            col = rgb(0,0,0,.25), 
            pch = 16, 
            cex = .5, 
            regLine = list(method = lm, 
                           lty = 1, 
                           lwd = 2, 
                           col = 1), 
            smooth = list(method = loess, 
                          spread = T, 
                          lty.smooth = 2, 
                          lwd.smooth = 2, 
                          col.smooth = "red",
                          lty.spread = 3, 
                          lwd.spread = 2, 
                          col.spread = "red"))
scatterplot(adm_8 ~ phexp1,  
            data = data_c, 
            boxplots = F, 
            grid = F, 
            jitter = list(x = 1), 
            col = rgb(0,0,0,.25), 
            pch = 16, 
            cex = .5, 
            regLine = list(method = lm, 
                           lty = 1, 
                           lwd = 2, 
                           col = 1), 
            smooth = list(method = loess, 
                          spread = T, 
                          lty.smooth = 2, 
                          lwd.smooth = 2, 
                          col.smooth = "red",
                          lty.spread = 3, 
                          lwd.spread = 2, 
                          col.spread = "red"))
scatterplot(adm_8 ~ vaxref_jan,  
            data = data_c, 
            boxplots = F, 
            grid = F, 
            jitter = list(x = 1), 
            col = rgb(0,0,0,.25), 
            pch = 16, 
            cex = .5, 
            regLine = list(method = lm, 
                           lty = 1, 
                           lwd = 2, 
                           col = 1), 
            smooth = list(method = loess, 
                          spread = T, 
                          lty.smooth = 2, 
                          lwd.smooth = 2, 
                          col.smooth = "red",
                          lty.spread = 3, 
                          lwd.spread = 2, 
                          col.spread = "red"))
scatterplot(adm_8 ~ nodoc,  
            data = data_c, 
            boxplots = F, 
            grid = F, 
            jitter = list(x = 1), 
            col = rgb(0,0,0,.25), 
            pch = 16, 
            cex = .5, 
            regLine = list(method = lm, 
                           lty = 1, 
                           lwd = 2, 
                           col = 1), 
            smooth = list(method = loess, 
                          spread = T, 
                          lty.smooth = 2, 
                          lwd.smooth = 2, 
                          col.smooth = "red",
                          lty.spread = 3, 
                          lwd.spread = 2, 
                          col.spread = "red"))
scatterplot(adm_8 ~ dmargin,  
            data = data_c, 
            boxplots = F, 
            grid = F, 
            jitter = list(x = 1), 
            col = rgb(0,0,0,.25), 
            pch = 16, 
            cex = .5, 
            regLine = list(method = lm, 
                           lty = 1, 
                           lwd = 2, 
                           col = 1), 
            smooth = list(method = loess, 
                          spread = T, 
                          lty.smooth = 2, 
                          lwd.smooth = 2, 
                          col.smooth = "red",
                          lty.spread = 3, 
                          lwd.spread = 2, 
                          col.spread = "red"))
scatterplot(adm_8 ~ density,  
            data = data_c, 
            boxplots = F, 
            grid = F, 
            jitter = list(x = 1), 
            col = rgb(0,0,0,.25), 
            pch = 16, 
            cex = .5, 
            regLine = list(method = lm, 
                           lty = 1, 
                           lwd = 2, 
                           col = 1), 
            smooth = list(method = loess, 
                          spread = T, 
                          lty.smooth = 2, 
                          lwd.smooth = 2, 
                          col.smooth = "red",
                          lty.spread = 3, 
                          lwd.spread = 2, 
                          col.spread = "red"))
scatterplot(adm_8 ~ poverty,  
            data = data_c, 
            boxplots = F, 
            grid = F, 
            jitter = list(x = 1), 
            col = rgb(0,0,0,.25), 
            pch = 16, 
            cex = .5, 
            regLine = list(method = lm, 
                           lty = 1, 
                           lwd = 2, 
                           col = 1), 
            smooth = list(method = loess, 
                          spread = T, 
                          lty.smooth = 2, 
                          lwd.smooth = 2, 
                          col.smooth = "red",
                          lty.spread = 3, 
                          lwd.spread = 2, 
                          col.spread = "red"))
scatterplot(adm_8 ~ plus65,  
            data = data_c, 
            boxplots = F, 
            grid = F, 
            jitter = list(x = 1), 
            col = rgb(0,0,0,.25), 
            pch = 16, 
            cex = .5, 
            regLine = list(method = lm, 
                           lty = 1, 
                           lwd = 2, 
                           col = 1), 
            smooth = list(method = loess, 
                          spread = T, 
                          lty.smooth = 2, 
                          lwd.smooth = 2, 
                          col.smooth = "red",
                          lty.spread = 3, 
                          lwd.spread = 2, 
                          col.spread = "red"))

```

Scatterplot of homoskedasticity
```{r}
plot(mod4, which = 1)
scatterplot(x = fitted(mod4),  
            y = resid(mod4), 
            main = "Residuals vs Fitted, model 4", 
            boxplots = F, 
            grid = F, 
            col = rgb(0,0,0,.5), 
            pch = 16, 
            cex = .5, 
            regLine = F, 
            smooth = list(method = loess, 
                          spread = T, 
                          lty.smooth = 2, 
                          lwd.smooth = 2, 
                          col.smooth = "red", 
                          lty.spread = 3, 
                          lwd.spread = 2, 
                          col.spread = "red"))

scatterplot(x = fitted(mod3),  
            y = abs(resid(mod3)), 
            main = "Absolute value of Residuals vs Fitted, model 4",
            boxplots = F, 
            grid = F, 
            col = rgb(0,0,0,.5), 
            pch = 16, 
            cex = .5, 
            regLine = list(method = lm, 
                           lty = 1, 
                           lwd = 2, 
                           col = 1), 
            smooth = list(method = loess, 
                          spread = T, 
                          lty.smooth = 2, 
                          lwd.smooth = 2, 
                          col.smooth = "red", 
                          lty.spread = 3, 
                          lwd.spread = 2, 
                          col.spread = "red")) 

bptest(mod3) #a Breusch-Pagan Test, we fail to reject the null hypothesis. We do not have sufficient evidence to say that heteroscedasticity is present in the regression model.(which means homoskedasticity)
```   
Normality of Residuals
```{r}
hist(resid(mod4), border = "white", main = "Histogram of model 4 residuals", breaks = 25) 
qqPlot(rstandard(mod4), 
       main = "Normal Q-Q, model 4", 
       col = rgb(0,0,0,.25), 
       pch = 16, 
       cex = .5, 
       col.lines = "red", 
       lwd = 1, 
       grid = F)
```
From 1st Table, 41 and 31 has higher phexp1, extremely lower density, no obvious pattern
Identified by cook's distance, Alaska is special with extremely phexp1 and lowest density; all 3 observations have low density
```{r}
# from qqplot and previous scatterplot
kable(data_c[c(36, 41, 31), c("adm_6", "exp", "phexp1", "vaxref_jan", "nodoc", "dmargin", "density", "poverty", "plus65")], "html")
plot(mod4, which = 4) 
plot(mod4, which = 5)
#from cook's distance
kable(data_c[c(2, 31, 41), c("adm_6", "exp", "phexp1", "vaxref_jan", "nodoc", "dmargin", "density", "poverty", "plus65")], "html")
data_c$cooksd <- cooks.distance(mod4)
#from rstudent
data_c$rstudent <- rstudent(mod4)
plot(data_c$rstudent,
     xlab = "Index",
     ylab = "Studentized residuals",
     pch = 19,
     main = "Distribution of Studentized Residuals")
abline(h=0, col = "red")
abline(h=2, col = "blue")
abline(h=-2, col = "blue")
abline(h=3, col = "green")
abline(h=-3, col = "green")
out1 <- subset(data_c, abs(data_c$rstudent) > 3 )
#c(41)
#from leverage
data_c$leverage <- hatvalues(mod4)
(2*9+2)/50
plot(data_c$leverage,
     xlab = "Index",
     ylab = "Leverage",
     pch = 19,
     main = "One Rule of Thumb")
abline(h=.4, col="blue")
out2 <- subset(data_c, abs(data_c$leverage) > 0.5 )
#c(2)
kable(data_c[c(2, 31, 36, 41, 45), c("state","adm_6", "exp", "phexp1", "vaxref_jan", "nodoc", "dmargin", "density", "poverty", "plus65")], "html")
```


Corrected model, AIC improves
```{r}
data_c$phexp1c = data_c$phexp1 - mean(data_c$phexp1)
data_c$phexp1c2 = data_c$phexp1c^2
mod4_5 <- rlm(adm_8 ~ exp + phexp1c2 + phexp1c + vaxref_jan + nodoc + dmargin + density + poverty + plus65 + mask_md, data=data_c)
stargazer(mod4, mod4_5,
          title = 'Table 5. Quadratic effect of public health expenditure', 
          ci=TRUE, ci.level = 0.95, 
          single.row =FALSE, 
          digits = 2, 
          no.space = TRUE, 
          type ='html',
          out = 'regression6.doc')

data_c$plus65c = data_c$plus65 - mean(data_c$plus65)
data_c$plus65c2 = data_c$plus65c^2
mod4_6 <- rlm(adm_8 ~ exp + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65c2 + plus65c + mask_md, data=data_c)
stargazer(mod4, mod4_6,
          title = 'Table 5. Quadratic effect of population above 65 years old', 
          ci=TRUE, ci.level = 0.95, 
          single.row =FALSE, 
          digits = 2, 
          no.space = TRUE, 
          type ='html',
          out = 'regression7.doc')

mod4_2 <- rlm(adm_8 ~ exp + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65 + mask_md, data = data_c) 
summary(mod4_2)
stargazer(mod4, mod4_2,  
          title = 'Table 5. Third model after correction', 
          ci=TRUE, ci.level = 0.95, 
          single.row =FALSE, 
          digits = 2, 
          no.space = TRUE, 
          type ='html',
          out = 'regression2.doc') 
AIC(mod4)/nobs(mod4)
AIC(mod4_2)/nobs(mod4_2)

mod2_2 <- rlm(adm_4 ~ exp + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65 + mask_md, data = data_c) 
mod3_2 <- rlm(adm_6 ~ exp + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65 + mask_md, data = data_c) 
stargazer(mod2_2, mod3_2, mod4_2, 
          title = 'Table 6. Corrected models',
          ci=TRUE, ci.level = 0.95,
          single.row = FALSE,
          digits = 2,
          no.space = TRUE,
          type = 'html',
          out = 'regression3.doc')
```

```{r}
#interaction effects
data_c$expc = data_c$exp - mean(data_c$exp)
mod1_4 <- rlm(adm_2 ~ expc + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65 + mask_md + expc : mask_md, data=data_c)

mod2_4 <- rlm(adm_4 ~ expc + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65 + mask_md + expc : mask_md, data=data_c)

mod3_4 <- rlm(adm_6 ~ expc + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65 + mask_md + expc : mask_md, data=data_c)

mod4_4 <- rlm(adm_8 ~ expc + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65 + mask_md + expc : mask_md, data=data_c)

mod5_4 <- rlm(adm_9 ~ expc + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65 + mask_md + expc : mask_md, data=data_c)


stargazer(mod2_4, mod3_4, mod4_4,
          title = 'Table 7. Interection effect of health expenditure', 
          ci=TRUE, ci.level = 0.95, 
          single.row =FALSE, 
          digits = 2, 
          no.space = TRUE, 
          type ='html',
          out = 'regression4.doc')
```

```{r}
#interaction effects 2
data_c$plus65c <- data_c$plus65 - mean(data_c$plus65)
mod1_3 <- rlm(adm_2 ~ exp + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65c + mask_md + plus65c:mask_md, data=data_c)

mod2_3 <- rlm(adm_4 ~ exp + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65c + mask_md + plus65c:mask_md, data=data_c)

mod3_3 <- rlm(adm_6 ~ exp + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65c + mask_md + plus65c:mask_md, data=data_c)

mod4_3 <- rlm(adm_8 ~ exp + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65c + mask_md + plus65c:mask_md, data=data_c)

mod5_3 <- rlm(adm_9 ~ exp + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65c + mask_md + plus65c:mask_md, data=data_c)


stargazer(mod2_3, mod3_3, mod4_3,
          title = 'Table 3. regression result', 
          ci=TRUE, ci.level = 0.95, 
          single.row =FALSE, 
          digits = 2, 
          no.space = TRUE, 
          type ='html',
          out = 'regression5.doc')
```

```{r}
anova(mod4, mod4_4)
anova(mod4, mod4_3)

stargazer(anova(mod4, mod4_4), 
          type = "html",
          out = 'anova1.doc')
stargazer(anova(mod4, mod4_3),
          type = 'html',
          out = 'anova2.doc')
#manually collected them into a table
```

```{r}
hist(data_month$feb_adm,breaks = 10, main = 'Figure 8.Histogram of feb_adm')
hist(data_month$april_adm,breaks = 10, main = 'Figure 8.Histogram of april_adm')
hist(data_month$june_adm,breaks = 10, main = 'Figure 8.Histogram of june_adm')
hist(data_month$august_adm,breaks = 10, main = 'Figure 8.Histogram of august_adm')
hist(data_month$sept_adm,breaks = 10, main = 'Figure 8.Histogram of sept_adm')
```

Another new model, dep var changes

```{r}
#Base Model
mod_feb <- lm(feb_adm ~ exp + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65 + mask_md, data=data_month)

mod_apr <- lm(april_adm ~ exp + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65 + mask_md, data=data_month)

mod_jun <- lm(june_adm ~ exp + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65 + mask_md, data=data_month)

mod_aug <- lm(august_adm ~ exp + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65 + mask_md, data=data_month)

mod_sep <- lm(sept_adm ~ exp + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65 + mask_md, data=data_month)

#Table
stargazer(mod_apr, mod_jun,mod_aug,
          title = 'Table 3. regression result', 
          ci=TRUE, ci.level = 0.95, 
          single.row =FALSE, 
          digits = 2, 
          no.space = TRUE, 
          type ='html',
          out = 'regression5.doc')

```


```{r}
#Models of interaction effects

data_month$expc = data_month$exp - mean(data_month$exp)
mod_feb_4 <- lm(feb_adm ~ expc + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65 + mask_md + expc : mask_md, data=data_month)

mod_apr_4 <- lm(april_adm ~ expc + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65 + mask_md + expc : mask_md, data=data_month)

mod_jun_4 <- lm(june_adm ~ expc + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65 + mask_md + expc : mask_md, data=data_month)

mod_aug_4 <- lm(august_adm ~ expc + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65 + mask_md + expc : mask_md, data=data_month)

mod_sep_4 <- lm(sept_adm ~ expc + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65 + mask_md + expc : mask_md, data=data_month)

#Table
stargazer(mod_apr_4, mod_jun_4,mod_aug_4,
          title = 'Table 3. regression result', 
          ci=TRUE, ci.level = 0.95, 
          single.row =FALSE, 
          digits = 2, 
          no.space = TRUE, 
          type ='html',
          out = 'regression6.doc')

```


```{r}
#Models of interaction effects
data_month$plus65c = data_month$plus65 - mean(data_month$plus65)
mod_feb_3 <- lm(feb_adm ~ exp + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65c + mask_md + plus65c : mask_md, data=data_month)

mod_apr_3 <- lm(april_adm ~ exp + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65c + mask_md + plus65c : mask_md, data=data_month)

mod_jun_3 <- lm(june_adm ~ exp + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65c + mask_md + plus65c : mask_md, data=data_month)

mod_aug_3 <- lm(august_adm ~ exp + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65c + mask_md + plus65c : mask_md, data=data_month)

mod_sep_3 <- lm(sept_adm ~ exp + phexp1 + vaxref_jan + nodoc + dmargin + density + poverty + plus65c + mask_md + plus65c : mask_md, data=data_month)

#Table
stargazer(mod_apr_3, mod_jun_3,mod_aug_3,
          title = 'Table 3. regression result', 
          ci=TRUE, ci.level = 0.95, 
          single.row =FALSE, 
          digits = 2, 
          no.space = TRUE, 
          type ='html',
          out = 'regression7.doc')

```


